---
title: "Tidy Data Lite: Exercises"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, error = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(dataAnim)
library(htmlwidgets)
library(shiny)
library(DSR)
```

## pivot_longer: One variable spread across multiple columns

- Column headers are values, not variables

**The pivot_longer funciton can help!**

```{r}
datoy_wide
```

- 3 variables
  + *Name* stored in rows
  + *Subject* spread across column names
  + *Score* stored in cell values
  
To tidy we need to use pivot_longer to make the following transformation

Run the following animation 

```{r, echo = FALSE}

 gather_anim(
     key = 'Subject',
     value = 'Score',
     data = datoy_wide,
     col = c("English","Maths")
 )
```


- The first arugment is the *data* to reshape *datoy_wide*
- The second argument describes which *cols* needs to be reshaped, in this case it's *English & Maths*
- The *names_to* argument gives the name of the variable that will be created from the data stored in the column names, i.e. *Subject*
- The *values_to* gives the name of the variable that will be created from the data stored in the cell value, i.e. *Score*
    + Neither the names_to nor the values_to column exists in datoy_wide, so we provide them as character strings surrounded in quotes.

Let's code it!

```{r}
?pivot_longer
datoy_wide

```


## You try!

The billboard dataset records the billboard rank of songs in the year 2000. It has a form similar to the relig_income data, but the data encoded in the column names is really a number, not a string.

Uncomment the code below and fill in to tidy the billboard data set 

```{r}
billboard %>% 
  head()

# billboard %>% 
#   pivot_longer(cols = , 
#                names_to = ,
#                names_prefix = "wk",
#                values_to = ,
#                values_drop_na = TRUE)
# 


```

What did

  - values_drop_na do ?
    + hint: not every song stays in the charts for all 76 weeks
  - names_preix do ? 

Which of the following would NOT have worked?

1. pivot_longer(data = billboard,cols = starts_with('wk'), names_to = "week", names_prefix = "wk", 
               values_to = "rank",values_drop_na = TRUE)
2. pivot_longer(data = billboard,cols = -c(artist:date.entered), names_to = "week", names_prefix = "wk", 
               values_to = "rank",values_drop_na = TRUE)
3. pivot_longer(data = billboard, cols = 4:79, names_to = "week", names_prefix = "wk", 
               values_to = "rank",values_drop_na = TRUE)
4. pivot_longer(data = billboard, cols = -artist:date.entered, names_to = "week", names_prefix = "wk", 
               values_to = "rank",values_drop_na = TRUE)
 
What does this add?

```{r}
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    names_prefix = "wk",
    names_ptypes = list(week = integer()),
    values_to = "rank",
    values_drop_na = TRUE,
  )
```

One more

- Tidy the preg data set below:

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

preg

# add your tidying code here 
```

## pivot_wider: An observation is scattered across multiple rows

- Some values are actually variables (i.e., variables stored in a column)
  + In this instance, we would use pivot_wider to make tidy data
  
```{r, echo = FALSE}
weather <- readr::read_csv("weather.csv") %>% 
  pivot_longer(d1:d31, names_to = "day", names_prefix = "d",
               values_to = "value", values_drop_na = TRUE) 
weather %>% 
  head()

```

- Most variables have individuals columns *(id, year,month, day, value)* but two variables
*(tmax, tmin)* are spread across rows, and stored in the "element" column.
  + Element is not a variable - it's a column that stores the names of two variables
  
- Fixing this requres **pivot_wider** function 
  + Performs the inverse of **pivot_longer** by spreading the *element* and *value* columns back out into columns
  
```{r}
weather %>% 
  pivot_wider(names_from = element, values_from = value) %>% 
  head()
```

- The first arugment is the *data* to reshape *weather*
- The second argument,*id_cols* describes which columns that unique identify each observation. The default is usually sufficient (defaults to all columns in the data except for the columns specified in names_from and values_from)
  + Typically only used when you have an additional variable that is directly related. Can be helpful in cleaning multiple choice data
- The *names_from* and *values_from* are paired arguments describing which column (or columns) to get the name of the output column *(name_from = element)*, and which column (or columns) to get the cell values from *(values_from = value)*
  + If values_from contains multiple values, the value will be added to the front of the output column. We will see an example of this in a second

- In practice, relatively rare to need pivot_wider for tidying, but it's often useful for creating summary tables for presentation

Remember we tidied the scores data 

```{r}
datoy_long
```

What if we wanted this in wide format for a summary table?

```{r}
datoy_long

spread_anim(
  key = "Subject",
  value = "Score", 
  data = datoy_long
)
```


Let's code it!

```{r}
datoy_long %>% 
  pivot_wider(names_from = Subject, values_from = Score)
```


## You try!

Tidy the us_rent_income data set 
  - contains information about median income and rent for each state in the US for 2017 (from the American Community Survey, retrieved with the tidycensus package).

Uncomment the code below and fill in to tidy the us_rent_income data set

Hint: both *estimte* and *moe* are value columns

```{r}
head(us_rent_income)
?us_rent_income

us_rent_income %>% 
  pivot_wider(
    names_from = , 
    values_from = c('estimate','moe')
  )

us_rent_income %>% 
  pivot_wider(
    names_from = variable, 
    values_from = c('estimate','moe')
  )
```

One More

Tidy "table2" using pivot_wider

```{r}
table2 %>% 
  head()

table2 %>% 
  pivot_wider(names_from = key, values_from = value)

```


## Putting it all together

Load & examine the following dataset 

```{r}
stocks <- tibble(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)

stocks
```

Make a new dataset "tidy_stocks" by using the appropriate pivot function to tidy

```{r}
# insert tidying code here

```

After you've tidied stocks, make another table "untidied_stocks" by putting "tidy_stocks"
back in its original format:

```{r}
head(stocks)
```


```{r}
#insert code that will convert your new "tidy_stocks" data set back to the "stocks" data set


```


## When things are a real mess!

- pivot_longer:

  + Many variables in column names
  + Multiple observations per row
  + Duplicate column names
  + Columns and rows are values of variables

- pivot_wider: 

  + filling in missing values
  + for aggregationg/summary table creation
  + column names from multiple variables
  + multi-choice data and id_cols

- pivot_longer + pivot_wider


- We will dive in after break!

## How am I doing?

Content | Presentation
------- | ------------
   +    |     +
   -    |     -
   

   
   
   
   
